`phyr`: An R package for phylogenetic species-distribution modelling in ecological communities

### 2019-10-06 ms.Rmd
add citations [skip ci]

Ecological communities are collections of species that occur within the same geographical area. Which species occur within communities depends on the dispersal ability of species to enter the community, the environmental conditions that they find there, and the interactions that they have with other species in the community. These three processes -- dispersal, environmental tolerance, and species interactions -- depend on the traits that species possess and hence reflect evolutionary history and biogeographic processes [@warren2014mistaking; @gerhold2018deep]. For example, an insect species might only occur in a lake if its adult stage has long-distance flight capabilities, if it can tolerate the low pH of the lake, and if it can avoid the predators that are common. Because traits play a central role in the composition of species that make up a community, community composition will likely reflect, at least in part, phylogenetic relationships among species. For example, two closely related insects might have similar dispersal capability, pH tolerance, and predator avoidance behavior, making them more likely to occur in the same lake. The recognition that phylogenetic relationships can increase our understanding of communities has led to a growing number of statistical methods for analyzing phylogenetic community composition.

Just as the distributions of two species might reflect their proximity on a phylogenetic tree, the species occurring at two sites might reflect the sites’ geographical proximity. The most immediate possible cause of spatial correlations in species distributions is dispersal, if nearby sites are more likely to be colonized by a species. Spatial proximity may also be a surrogate for environmental variables that are unknown or unmeasured. For example, an insect species might occur in two nearby lakes because they both have low pH, yet pH has not been measured. Just as phylogenetic relationships among species can generate correlations between species in which sites they occupy, so too can spatial proximity generate correlations between sites in the species they contain.

@@

Statistical models for phylogenetic community composition provide flexible tools for exploring the many possible factors underlying the distribution of species and the composition of communities. The models can describe complex relationships in the data, such as how phylogenetically related species might respond similarly to the same environmental gradient, or how phylogenetically related species might exclude each other from the same communities. They also give a firm statistical basis to test these patterns, the ability to simulate data sets from the fitted model, and the ability to predict the composition of unsurveyed communities. These benefits of phylogenetic community composition models come with costs: building models can be intricate and fitting them computationally slow. The R package `phyr` is designed to overcome many of the costs with a user-friendly interface, flexibility to build a rich collection of models, and good computational performance. Below, we first give a brief overview of the structure and syntax of two key functions `pglmm()` and `cor_phylo()`. We then compare them to methods and programs that are currently available. Finally, we give applications of `pglmm()` and `cor_phylo()` to simulated data to illustrate their implementation and output.

@@

`Phyr` contains three groups of functions (Table \@ref(tab:mainFunc)): phylogenetic generalized linear mixed models (`pglmm()`), phylogenetic correlations between functional traits (`cor_phylo()`), and phylogenetic diversity metrics (e.g. `psv()`, `pse()`). The workhorse functions of all groups are written in C++ to increase computational speed. Here, we will focus on the first two groups of functions because they are more complicated than phylogenetic diversity metrics.

@@

`pglmm()` constructs and fits generalized linear mixed models that incorporate covariance matrices containing the phylogenetic relationships among species. The syntax for `pglmm()` resembles that used in the R package `lme4`, and indeed `pglmm()` will fit most of the models that can be fit with `lmer()` and `glmer()`. `pglmm()` goes beyond `lmer()`and `glmer()` by allowing the specification of covariance matrices, which could be a phylogenetic covariance matrix or any other covariance matrix that the user defines (e.g., spatial or temporal autocorrelation matrix). `pglmm()` can also fit models with “nested” covariance structures (e.g. species phylogenetic covariance matrix nested within site covariance matrix). `pglmm()` can operate in both frequentist mode, with the distribution of species among communities being Gaussian, Binary, Binomial or Poisson, and Bayesian mode with the addition of zero-inflated Binomial and Poisson distributions. Finally, it is our hope that the formula syntax of `pglmm()` can be used to fit similar models with other programs such as Stan (e.g. via R package `brm`).

@@

Here, Y is a binary (Bernoulli) dependent variable which takes values of either 0 or 1. The specification `family = 'binomial'` allows binary data and also binomial data for which Y is a matrix containing columns for successes and failures. The independent variables `trait` and `env` take on different values for each species and site, respectively. Sites and species (`sp`) are treated as random effects: thus, `(1|site)` implies that a value from a Gaussian random variable is picked for each site, thereby representing unmeasured differences among sites. For the case of species, the double underscore in `(1|sp__)` implies that, in addition to a random effect for species, there is a second random effect which contains the phylogenetic relationships among species. The phylogenetic random effect assumes that values for each species are picked from a multivariate Gaussian distribution with phylogenetic covariance matrix 𝚺. A covariance matrix 𝚺 is specified by `cov_ranef = list(sp = phy.sp, site = V.space)`. The covariance matrix `phy.sp` associated with species can be a `phylo` object from the R package `ape`. It is also possible to specify an explicit covariance matrix, such as `site = V.space`. To construct 𝚺 from a 'phylo' object, `pglmm()` assumes that the residual variation associated with species follows a Brownian motion model of evolution, so that the covariance between two species is proportional to the distance between them on the phylogenetic tree.

The syntax `(1|sp__)` or  `(1|site__)` generates two random effects, one without and one with phylogenetic or spatial covariances; in contrast, `(1|sp)` would generate only a single random effect that is independent among species. `pglmm()` forces in a term for `(1|sp)` whenever `(1|sp__)` is specified, because otherwise any difference among species would be captured by the diagonal elements in 𝚺 even in the absence of covariances among phylogenetically related species which are specified by the off-diagonal elements of 𝚺. Therefore, if `(1|sp)` were not included, this could lead to the identification of phylogenetic signal in the abundances of species even in its absence. To account for differences among sites in how they select for species with different traits, `(trait|site)` allows the slope of Y against `trait` to be a Gaussian random variable. Similarly, to account for the differences among species for how they respond to `env`, `(env|sp__)` allows the relationship of Y against `env` to be given by two slopes, the first slope that is picked from a Gaussian random variable in which species are independent and the second slope that is picked from a multivariate Gaussian with covariance matrix 𝚺. Finally, `(1|sp__@site)` generates a nested term: within a site, the residual variation in Y shows phylogenetic relatedness, with phylogenetically related species more likely to occur in the same site. Note that `(1|sp__)` differs from `(1|sp__@site)` because `(1|sp__)` generates differences in the mean value of Y for species across all sites, whereas `(1|sp__@site)` is local, giving the covariances among species only within sites. Other forms of a nested term are available in `pglmm()`, which can be used to study more complicated questions such as bipartite networks.

With `bayes = FALSE`, `pglmm()` is fit using a frequentist approach. ML or REML is used for fitting, with `REML = TRUE` as the default. For a non-Gaussian model (e.g., `family = 'binomial'`), an iterated quasi-likelihood method is used for model fitting which gives the approximate likelihood; p-values for the fixed effects are given by a Wald test and for the random effects by profile likelihood, although we recommend bootstrap-based tests when computationally feasible. Note that `REML = TRUE` is an option for non-Gaussian models (in contrast to `glmer`) due to the algorithm used. With `bayes = TRUE`, a Bayesian approach is implemented using INLA [@rue2009approximate], which gives parameter estimates and credible intervals. For large problems with the number of species-site combinations exceeding 2000, the Bayesian computations are considerably faster than the frequentist computations. Finally, a key to interpreting the results from a model is understanding the structure of the covariance matrices associated with the random effects. Therefore, `pglmm()` has associated plotting functions `pglmm.plot.ranef()` that present the design matrices for the random effects (Fig. \@ref(fig:designPlot)).

@@

`pglmm()` and `cor_phylo()` both appear in existing R packages (`pez` and `ape`, respectively), although the versions in `phyr` represent considerable improvements in ease-of-use, computational speed, and flexibility. Both have new syntax that makes them more intuitive to use. `pglmm()` also has new associated functions that plot the design of the covariance matrices (Fig. \@ref(fig:designPlot)), making model interpretation easier. Both are now coded in C++ (for key functions), which speeds computation time by 5-10X. `pglmm()` now supports several non-Gaussian distributions and allows Bayesian analyses using INLA that is particularly useful for large datasets. Finally, both include more output; for example, both now include facilities to perform likelihood ratio tests and compute AIC and BIC values for model comparisons.

>>>>

Ecological communities are collections of species that occur within the same geographical area. Which species occur within communities depends on the dispersal ability of species to enter the community, the environmental conditions that they find there, and the interactions that they have with other species in the community. These three processes -- dispersal, environmental tolerance, and species interactions -- depend on the traits that species possess and hence reflect evolutionary history and biogeographic processes [@warren2014mistaking; @gerhold2018deep]. For example, an insect species might only occur in a lake if its adult stage has long-distance flight capabilities, if it can tolerate the low pH of the lake, and if it can avoid the predators that are common. Because traits play a central role in the composition of species that make up a community, community composition will likely reflect, at least in part, phylogenetic relationships among species. For example, two closely related insects might have similar dispersal capability, pH tolerance, and predator avoidance behavior, making them more likely to occur in the same lake. The recognition that phylogenetic relationships can increase our understanding of communities has led to a growing number of statistical methods for analyzing phylogenetic community composition [@ives2011generalized; @li2017can; @frishkoff2017phylogenetic].

Just as the distributions of two species might reflect their proximity on a phylogenetic tree, the species occurring at two sites might reflect the sites’ geographical proximity. The most immediate possible cause of spatial correlations in species distributions is dispersal, if nearby sites are more likely to be colonized by a species. Spatial proximity may also be a surrogate for environmental variables that are unknown or unmeasured. For example, an insect species might occur in two nearby lakes because they both have low pH, yet pH has not been measured. Just as phylogenetic relationships among species can generate correlations between species in which sites they occupy, so too can spatial proximity generate correlations between sites in the species they contain [@ives2006statistics].

@@

Statistical models for phylogenetic community composition provide flexible tools for exploring the many possible factors underlying the distribution of species and the composition of communities [@ives2011generalized]. The models can describe complex relationships in the data, such as how phylogenetically related species might respond similarly to the same environmental gradient, or how phylogenetically related species might exclude each other from the same communities. They also give a firm statistical basis to test these patterns, the ability to simulate data sets from the fitted model, and the ability to predict the composition of unsurveyed communities. These benefits of phylogenetic community composition models come with costs: building models can be intricate and fitting them computationally slow. The R package `phyr` is designed to overcome many of the costs with a user-friendly interface, flexibility to build a rich collection of models, and good computational performance. Below, we first give a brief overview of the structure and syntax of two key functions `pglmm()` and `cor_phylo()`. We then compare them to methods and programs that are currently available. Finally, we applied `pglmm()` and `cor_phylo()` to simulated data to illustrate their implementation and output.

@@

`Phyr` contains three groups of functions (Table \@ref(tab:mainFunc)): phylogenetic generalized linear mixed models (`pglmm()`), phylogenetic correlations between functional traits (`cor_phylo()`), and community phylogenetic diversity metrics (e.g. `psv()`, `pse()`). The workhorse functions of all groups are written in C++ to increase computational speed. Here, we will focus on the first two groups of functions because they are more complicated than community phylogenetic diversity metrics.

@@

`pglmm()` constructs and fits generalized linear mixed models that incorporate covariance matrices containing the phylogenetic relationships among species. The syntax for `pglmm()` resembles that used in the R package `lme4` [@bates2015lme4], and indeed `pglmm()` will fit most of the models that can be fit with `lmer()` and `glmer()`. `pglmm()` goes beyond `lmer()`and `glmer()` by allowing the specification of covariance matrices, which could be a phylogenetic covariance matrix or any other covariance matrix that the user defines (e.g., spatial or temporal autocorrelation matrix). `pglmm()` can also fit models with “nested” covariance structures (e.g. species phylogenetic covariance matrix nested within site covariance matrix). `pglmm()` can operate in both frequentist mode, with the distribution of species among communities being Gaussian, Binary, Binomial or Poisson, and Bayesian mode with the addition of zero-inflated Binomial and Poisson distributions. Finally, it is our hope that the formula syntax of `pglmm()` can be used to fit similar models with other programs such as Stan (e.g. via R package `brm` [@brmsR2018]).

@@

Here, Y is a binary (Bernoulli) dependent variable which takes values of either 0 or 1. The specification `family = 'binomial'` allows binary data and also binomial data for which Y is a matrix containing columns for successes and failures. The independent variables `trait` and `env` take on different values for each species and site, respectively. Sites and species (`sp`) are treated as random effects: thus, `(1|site)` implies that a value from a Gaussian random variable is picked for each site, thereby representing unmeasured differences among sites. For the case of species, the double underscore in `(1|sp__)` implies that, in addition to a random effect for species, there is a second random effect which contains the phylogenetic relationships among species. The phylogenetic random effect assumes that values for each species are picked from a multivariate Gaussian distribution with phylogenetic covariance matrix 𝚺. A covariance matrix 𝚺 is specified by `cov_ranef = list(sp = phy.sp, site = V.space)`. The covariance matrix `phy.sp` associated with species can be a `phylo` object from the R package `ape` [@ape2018]. It is also possible to specify an explicit covariance matrix, such as `site = V.space`. To construct 𝚺 from a 'phylo' object, `pglmm()` assumes that the residual variation associated with species follows a Brownian motion model of evolution, so that the covariance between two species is proportional to the distance between them on the phylogenetic tree.

The syntax `(1|sp__)` or  `(1|site__)` generates two random effects, one without and one with phylogenetic or spatial covariances; in contrast, `(1|sp)` would generate only a single random effect that is independent among species. `pglmm()` forces in a term for `(1|sp)` whenever `(1|sp__)` is specified, because otherwise any difference among species would be captured by the diagonal elements in 𝚺 even in the absence of covariances among phylogenetically related species which are specified by the off-diagonal elements of 𝚺. Therefore, if `(1|sp)` were not included, this could lead to the identification of phylogenetic signal in the abundances of species even in its absence. To account for differences among sites in how they select for species with different traits, `(trait|site)` allows the slope of Y against `trait` to be a Gaussian random variable. Similarly, to account for the differences among species for how they respond to `env`, `(env|sp__)` allows the relationship of Y against `env` to be given by two slopes, the first slope that is picked from a Gaussian random variable in which species are independent and the second slope that is picked from a multivariate Gaussian with covariance matrix 𝚺. Finally, `(1|sp__@site)` generates a nested term: within a site, the residual variation in Y shows phylogenetic relatedness, with phylogenetically related species more likely to occur in the same site. Note that `(1|sp__)` differs from `(1|sp__@site)` because `(1|sp__)` generates differences in the mean value of Y for species across all sites, whereas `(1|sp__@site)` is local, giving the covariances among species only within sites. This nested term can be used to test for community clustering or overdispersion [@webb2002phylogenies; @ives2011generalized]. Other forms of a nested term are available in `pglmm()`, which can be used to study more complicated questions such as bipartite networks.

With `bayes = FALSE`, `pglmm()` is fitted using a frequentist approach. ML or REML is used for fitting, with `REML = TRUE` as the default. For a non-Gaussian model (e.g., `family = 'binomial'`), an iterated quasi-likelihood method is used for model fitting which gives the approximate likelihood; p-values for the fixed effects are given by a Wald test and for the random effects by profile likelihood, although we recommend bootstrap-based tests when computationally feasible. Note that `REML = TRUE` is an option for non-Gaussian models (in contrast to `glmer`) due to the algorithm used. With `bayes = TRUE`, a Bayesian approach is implemented using INLA [@rue2009approximate], which gives parameter estimates and credible intervals. For large problems with the number of species-site combinations exceeding 2000, the Bayesian computations are considerably faster than the frequentist computations. Finally, a key to interpreting the results from a model is understanding the structure of the covariance matrices associated with the random effects. Therefore, `pglmm()` has associated plotting functions `pglmm.plot.ranef()` that present the design matrices for the random effects (Fig. \@ref(fig:designPlot)).

@@

Earlier versions of `pglmm()` and `cor_phylo()` both appear in existing R packages (`pez` [@pearse2015pez] and `ape` [@ape2018], respectively), although the versions in `phyr` represent considerable improvements in ease-of-use, computational speed, and flexibility. Both have new syntax that makes them more intuitive to use. `pglmm()` also has new associated functions that plot the design of the covariance matrices (Fig. \@ref(fig:designPlot)), making model interpretation easier. Both are now coded in C++ (for key functions), which speeds computation time by 5-10X. `pglmm()` now supports several non-Gaussian distributions and allows Bayesian analyses using INLA that is particularly useful for large datasets. Finally, both include more output; for example, both now include facilities to perform likelihood ratio tests and compute AIC and BIC values for model comparisons.
